---
title: "hwset1"
author: "Katie Daisey"
date: "Monday, February 09, 2015"
output: html_document
---

#Homework Set 1

##Question 1
Two analytical methods (XRF and ICP) were applied to random areas on the same semiconductor material with a trace sodium contaminant.  The following data was obtained:
```{r echo=FALSE}
xrf<-c(85.1, 81.4, 77.1, 84.5, 87.9, 83.2, 86.6, 83.9, 81.1, 80.8)
    icp<-c(87.4, 90.1, 86.2, 89.2, 88.4, 82.9, 81.9, 87.4, 82.1, 80.6)
    data.1<-cbind(xrf,icp)
```
```{r}
data.1
```

a) Treating these methods as separate, but producing replicated results, the results _ distinguishable at 95% confidence and _ at 90% confidence.

b) If we instead treat the data as pairs of non-replicated samples, they are _ at 95% confidence and _ at 90% confidence.

c) what p values mean

d) 95% vs 90% confidence
e) more statistical power



##Question 2
Rutherford and Geiger (Phil. Mag. (1910)20, 698-707) counted alpha particles emitted by polonium using scintillation.  With N as the number of particles and f as the frequency N particles were observed during fixed time intervals, the following data was reported:
```{r, echo=4, eval=TRUE}
N<-c(0:14)
f<-c(57,203,383,525,532,408,273,139,45,27,10,4,0,1,1)
gold<-cbind(N,f)
gold
```
a) The mean number of alpha particles emitted in the fixed time interval can be calculated by finding the weighted mean, ie multiplying each N value by the corresponding f value and dividing by the total number of observations:
```{r}
particles<-N*f
total_particles<-sum(particles)
total_obs<-sum(f)
weighted_mean<-signif(total_particles/total_obs,3)
```
giving a mean number of particles of `r weighted_mean`.

b)A Poisson distribution


##Question 3
a) 
